---
title: "Figure 5"
editor_options: 
  chunk_output_type: console
format: gfm
execute:
  echo: true
  warning: false
  message: false
  error: false
---

```{r libraries, echo=FALSE}
library(tidyverse)
library(patchwork)
library(Seurat)
```

```{r setup}
#| echo: false

# Set a global theme and base size
theme_set(
  theme_minimal()
)

umap_min_text_size <- 8
umap_geom_text_size <- umap_min_text_size / ggplot2::.pt

update_geom_defaults("text", list(size = umap_geom_text_size, family = "Arial"))

umap_arial_text_min_8 <- theme(
  text = element_text(family = "Arial", size = 8),
  axis.text = element_text(family = "Arial", size = 8),
  axis.title = element_text(family = "Arial", size = 8),
  strip.text = element_text(family = "Arial", size = 8),
  legend.text = element_text(family = "Arial", size = 8),
  legend.title = element_text(family = "Arial", size = 8),
  plot.title = element_text(family = "Arial", size = 8)
)

fig_5_dir <- "fig_5_files"
dir.create(fig_5_dir, showWarnings = FALSE, recursive = TRUE)

figure5_pbmc_dir <- here::here("data", "processed", "02_internal_pbmc_scrna")
figure5_pbmc_intermediate_dir <- here::here("data", "intermediate", "02_internal_pbmc_scrna")
dir.create(figure5_pbmc_intermediate_dir, showWarnings = FALSE, recursive = TRUE)

figure5_umap_object_path <- here::here(
  "data", "processed", "02_internal_pbmc_scrna",
  "seurat.integrated2_monaco_umap_res0.9.rds"
)
figure5_pseudobulk_counts_path <- here::here(
  "data", "intermediate", "02_internal_pbmc_scrna",
  "figure5_pseudobulk_counts_by_cell_type_sample.rds"
)
figure5_de_time_path <- here::here(
  "data", "processed", "02_internal_pbmc_scrna",
  "figure5_time_de_by_cell_type.rds"
)

```

```{r}
load(here::here("data", "processed", "01_proteomics_metabolomics", "Data.RData"))
```

# Flow

```{r}
load(here::here("data", "intermediate", "flow_gating", "bcell.RData"))
load(here::here("data", "intermediate", "flow_gating", "tcell.RData"))
load(here::here("data", "intermediate", "flow_gating", "monocyte.RData"))
load(here::here("data", "intermediate", "flow_gating", "dcnk.RData"))

flow = list()
flow$panelResults = list(bcell=bcell,dcnk=dcnk,monocyte=monocyte,tcell=tcell)
rm(bcell,dcnk,monocyte,tcell)

flow$propsLong = do.call(rbind,lapply(flow$panelResults,function(x){x$propsLong}))
flow$propsWide = pivot_wider(flow$propsLong,id_cols = id,names_from = cluster,values_from = freq)%>%column_to_rownames("id")
flow$ncells = pivot_wider(flow$propsLong,id_cols = id,names_from = cluster,values_from = n)%>%column_to_rownames("id")
flow$total_cells = pivot_wider(flow$propsLong,id_cols = id,names_from = cluster,values_from = total)%>%column_to_rownames("id")

has_sample_data = gsub("_.*","",rownames(flow$propsWide))%in%rownames(data$sampleData)

flow$propsLong = flow$propsLong[flow$propsLong$id%in%rownames(flow$propsWide)[has_sample_data],]
flow$propsWide  = flow$propsWide[has_sample_data,]
flow$ncells = flow$ncells[has_sample_data,]
flow$total_cells = flow$total_cells[has_sample_data,]


flow_sample_ids <- gsub("_.*", "", rownames(flow$propsWide))
sample_aligned_names <- names(data)[vapply(data, function(x) {
  (is.data.frame(x) || is.matrix(x)) &&
    !is.null(rownames(x)) &&
    all(flow_sample_ids %in% rownames(x))
}, logical(1))]
flow$data = lapply(data[sample_aligned_names],function(x){
  x[flow_sample_ids,,drop = FALSE]
})

flow$panel = gsub("\\..*","",rownames(flow$propsLong))[match(colnames(flow$propsWide),flow$propsLong$cluster)]
flow$panel = c("T Cell","B Cell","DCNK","Monocyte")[match(flow$panel,c("tcell","bcell","dcnk","monocyte"))]
flow$panel[9:11] = "NK"
flow$panel[12:15] = "DC"
flow$propsLong$cluster = gsub(" NK Cells","",flow$propsLong$cluster)


flow$cluster = colnames(flow$propsWide)
colnames(flow$propsWide) = paste0(flow$panel," - ",colnames(flow$propsWide))
colnames(flow$propsWide) = gsub("yd","gd",colnames(flow$propsWide))
colnames(flow$propsWide) = gsub(" NK Cells","",colnames(flow$propsWide))
flow$propsLong$cluster = gsub("yd","gd",flow$propsLong$cluster)
```

# 5a
### PCA patients and controls all timepoints

### Panel A: Flow Cytometry PCA Across Cohorts
```{r}
# sams = flow$data$sampleData$Condition=="Patient"&
#   flow$data$sampleData$days_of_prior_antibiotics==0

## one joint PCA, then facet
sd <- flow$data$sampleData

keep <- sd$days_of_prior_antibiotics == 0 &
  sd$Condition %in% c("Patient","Control") &
  !grepl("35", sd$Site_s_of_EM)

d <- flow$propsWide[, !grepl("Dump|Debris|DP |DN |- DN|- DP|UNKN|IL1B", colnames(flow$propsWide))]
d <- apply(d, 2, function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))

keep <- keep & apply(d, 1, function(x) sum(is.na(x)) < 12) & !grepl("_1", rownames(d))
d <- d[keep, , drop = FALSE]

## impute once
d2 <- d
colnames(d2) <- make.names(colnames(d2))
m <- mice::mice(d2, 1, seed = 123, printFlag = FALSE)
d_imp <- mice::complete(m, 1)
colnames(d_imp) <- colnames(d)

## PCA once
pca <- prcomp(d_imp, scale. = FALSE)

vars <- paste0(
  "PC", 1:ncol(pca$x),
  " (", round(pca$sdev^2 / sum(pca$sdev^2) * 100, 1), "%)"
)

## build plotting df (no dplyr needed)
df <- data.frame(sample_id = rownames(pca$x), pca$x, stringsAsFactors = FALSE)
df$id <- sub(" .*", "", df$sample_id)
df$Condition <- sd[df$sample_id, "Condition"]
df$time <- sd[df$sample_id, "time"]
df$dataset <- ifelse(df$Condition == "Patient", "Patients", "Controls")

## optional: mix id levels so hues do not clump by facet
set.seed(1)
df$id <- factor(df$id, levels = sample(sort(unique(df$id))))

fig_5a <- ggplot(df, aes(PC1, PC2, color = id, shape = time)) +
  geom_point(size = 8, alpha = 0.5) +
  facet_wrap(~dataset) +
  scale_color_discrete(guide = "none") +
  scale_shape_manual(values = c("T1" = 15, "T2" = 16, "T3" = 17, "T4" = 18)) +
  xlab(vars[1]) + ylab(vars[2]) +
  theme_minimal(base_size = 50)


ggsave(
      plot = fig_5a,
      filename = file.path(fig_5_dir, "figure_05a_flow_cytometry_pca.png"),
      device = ragg::agg_png,
      width = 18,
      height = 9,
      units = "in",
      dpi = 600
    )

fig_5a


```


# 5b
#### Selected Boxplots

### Panel B: Plasmablast Abundance Over Time
```{r}
pop = 'B Cell - Plasmablasts'
sample_ids_5b <- rownames(flow$data$sampleData)
sams = (
  flow$data$sampleData$days_of_prior_antibiotics == 0 &
    flow$data$sampleData$Condition == "Patient" &
    as.character(flow$data$sampleData$Subject_ID) != "201455" &
    !sample_ids_5b %in% c("201455 T1", "201455 T2")
)
sd = flow$data$sampleData[sams,]
dat = data.frame(flow$propsWide[sams,pop],sd)
colnames(dat)[1] = pop
dat$time_numeric = as.numeric(gsub("^T(.)","\\1",dat$time))

fig_5b <- ggplot(dat,aes(x=time,y=`B Cell - Plasmablasts`))+
  geom_boxplot(outlier.shape = NA, size = 0.1)+
  geom_point(alpha=0.25,size=0.5)+
  geom_line(aes(group=Subject_ID),alpha=0.3, size = 0.1)+
  theme_minimal(base_size = 8)+
  ggtitle("Plasmablasts")+
  xlab("")+ylab("Proportion of B Cells")
fig_5b
ggsave(
  file.path(fig_5_dir, "figure_05b_paired_plasmablast_abundance.png"),
  fig_5b,
  device = ragg::agg_png,
  width = 2.25,
  height = 1.5,
  units = "in",
  dpi = 300
)
```

# 5c
#### scRNAseq
```{r load_scrna}
res_0.9 <- readRDS(figure5_umap_object_path)
DefaultAssay(res_0.9) <- "originalexp"

figure5_cell_type_full_levels <- c(
  "Progenitor cells",
  "Classical monocytes",
  "Intermediate monocytes",
  "Non classical monocytes",
  "Myeloid dendritic cells",
  "Plasmacytoid dendritic cells",
  "Natural killer cells",
  "Naive B cells",
  "Non-switched memory B cells",
  "Switched memory B cells",
  "Exhausted B cells",
  "Plasmablasts",
  "Naive CD4 T cells",
  "T regulatory cells",
  "Th1 cells",
  "Th2 cells",
  "Th17 cells",
  "Follicular helper T cells",
  "Th1/Th17 cells",
  "Terminal effector CD4 T cells",
  "Naive CD8 T cells",
  "Effector memory CD8 T cells",
  "Central memory CD8 T cells",
  "Terminal effector CD8 T cells",
  "MAIT cells",
  "Non-Vd2 gd T cells",
  "Vd2 gd T cells",
  "Low-density basophils"
)

figure5_cell_type_short_to_full <- c(
  "Progenitor" = "Progenitor cells",
  "Classical mono" = "Classical monocytes",
  "Int mono" = "Intermediate monocytes",
  "Non-class mono" = "Non classical monocytes",
  "Myeloid DCs" = "Myeloid dendritic cells",
  "pDCs" = "Plasmacytoid dendritic cells",
  "NK cells" = "Natural killer cells",
  "Naive B" = "Naive B cells",
  "Non-switch mem B" = "Non-switched memory B cells",
  "Switched mem B" = "Switched memory B cells",
  "Exhausted B" = "Exhausted B cells",
  "Plasmablasts" = "Plasmablasts",
  "Naive CD4 T" = "Naive CD4 T cells",
  "Treg" = "T regulatory cells",
  "Th1" = "Th1 cells",
  "Th2" = "Th2 cells",
  "Th17" = "Th17 cells",
  "Tfh" = "Follicular helper T cells",
  "Th1/Th17" = "Th1/Th17 cells",
  "Temra CD4 T" = "Terminal effector CD4 T cells",
  "Naive CD8 T" = "Naive CD8 T cells",
  "Tem CD8 T" = "Effector memory CD8 T cells",
  "Tcm CD8 T" = "Central memory CD8 T cells",
  "Temra CD8 T" = "Terminal effector CD8 T cells",
  "MAIT" = "MAIT cells",
  "Non-Vd2 gd T" = "Non-Vd2 gd T cells",
  "Vd2 gd T" = "Vd2 gd T cells",
  "LD basophils" = "Low-density basophils"
)

res_0.9@meta.data <- res_0.9@meta.data |>
  mutate(
    cell_type_fine_full = recode(
      as.character(cell_type_fine),
      !!!figure5_cell_type_short_to_full
    ),
    cell_type_fine_full = factor(
      cell_type_fine_full,
      levels = figure5_cell_type_full_levels
    )
  )
```

### Panel 5C: PBMC UMAP by Condition and Time

```{r fig_5c}
fig_5c_cell_type_levels <- c(
  "Classical mono",
  "Exhausted B",
  "Int mono",
  "MAIT",
  "Myeloid DCs",
  "Naive B",
  "Naive CD4 T",
  "Naive CD8 T",
  "NK cells",
  "Non-class mono",
  "Non-switch mem B",
  "Non-Vd2 gd T",
  "pDCs",
  "Plasmablasts",
  "Progenitor",
  "Switched mem B",
  "Tcm CD8 T",
  "Tem CD8 T",
  "Temra CD4 T",
  "Temra CD8 T",
  "Tfh",
  "Th1",
  "Th1/Th17",
  "Th17",
  "Th2",
  "Treg",
  "Vd2 gd T"
)

fig_5c_cell_type_colors <- scales::hue_pal()(length(fig_5c_cell_type_levels)) |>
  setNames(fig_5c_cell_type_levels)

fig_5c_cell_type_color_labels <- sprintf(
  "<span style='color:%s'>%s</span>",
  fig_5c_cell_type_colors,
  names(fig_5c_cell_type_colors)
) |>
  setNames(names(fig_5c_cell_type_colors))

fig_5c_subset_cell_type_levels <- c(
  "Non-switch mem B",
  "Naive B",
  "Naive CD8 T",
  "Tfh",
  "Th1",
  "Naive CD4 T",
  "Classical mono",
  "Vd2 gd T",
  "Th1/Th17",
  "MAIT",
  "Non-Vd2 gd T",
  "Myeloid DCs"
)

fig_5c_legend_cell_type_levels <- sort(fig_5c_subset_cell_type_levels)

condition_time_levels <- c(
  "Control",
  "Patient_T1",
  "Patient_T4"
)

condition_time_labels <- str_replace_all(
  condition_time_levels,
  "_",
  " "
)

set.seed(111)

Fig5C_df <- res_0.9@meta.data |>
  base::cbind(res_0.9@reductions$umap@cell.embeddings)

Fig5C_df_subset <- Fig5C_df |>
  filter(cell_type_fine %in% fig_5c_subset_cell_type_levels)

fig_5c_target_n <- Fig5C_df_subset |>
  filter(sample %in% c("L211932_T1", "L214612_T1") | Condition == "Patient") |>
  count(Condition, time) |>
  pull(n) |>
  min()

fig_5c_scaling_factors <- Fig5C_df_subset |>
  filter(sample %in% c("L211932_T1", "L214612_T1") | Condition == "Patient") |>
  count(Condition, time) |>
  mutate(scaling_factor = fig_5c_target_n / n)

Fig5C_df <- Fig5C_df_subset |>
  filter(sample %in% c("L211932_T1", "L214612_T1") | Condition == "Patient") |>
  group_by(Condition, time, cell_type_fine) |>
  group_modify(~ {
    scale_val <- fig_5c_scaling_factors |>
      filter(Condition == .y$Condition, time == .y$time) |>
      pull(scaling_factor)

    n_to_sample <- round(nrow(.x) * scale_val)

    slice_sample(.x, n = n_to_sample)
  }) |>
  ungroup() |>
  mutate(
    cell_type_fine = str_replace_all(cell_type_fine, " cells", ""),
    condition_time = factor(
      str_c(Condition, time, sep = "_") |>
        str_replace("Control.*", "Control") |>
        str_replace_all("_", " "),
      levels = condition_time_labels
    ),
    cell_type_fine = factor(cell_type_fine, levels = fig_5c_cell_type_levels)
  )

fig_5c <- ggplot(Fig5C_df, aes(umap_2, umap_1, color = cell_type_fine)) +
  geom_point(
    size = 0.3,
    stroke = 0,
    shape = 19,
    key_glyph = ggplot2::draw_key_blank
  ) +
  facet_wrap(~ condition_time) +
  coord_equal() +
  scale_color_manual(
    values = fig_5c_cell_type_colors,
    breaks = fig_5c_legend_cell_type_levels,
    labels = fig_5c_cell_type_color_labels[fig_5c_legend_cell_type_levels]
  ) +
  ggthemes::theme_tufte(base_size = 8, base_family = "Arial") +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  guides(
    color = guide_legend(
      title = NULL,
      ncol = 4,
      byrow = TRUE,
      keywidth = unit(0, "pt"),
      keyheight = unit(7, "pt"),
      override.aes = list(size = 0, alpha = 0)
    )
  ) +
  umap_arial_text_min_8 +
  theme(
    legend.position = "bottom",
    legend.location = "panel",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.title = element_blank(),
    legend.text = ggtext::element_markdown(
      family = "Arial",
      size = 8,
      margin = margin(0, 0, 0, 0)
    ),
    legend.key = element_blank(),
    legend.key.width = unit(0, "pt"),
    legend.key.height = unit(7, "pt"),
    legend.spacing.x = unit(3, "pt"),
    legend.spacing.y = unit(0, "pt"),
    legend.margin = margin(t = 2, r = 0, b = 0, l = 0),
    legend.box.margin = margin(0, 0, 0, 0),
    legend.box.spacing = unit(0, "pt"),
    plot.margin = margin(2, 2, 0, 2)
  )

ggsave(
  here::here("fig_5_files", "figure_05c_pbmc_umap_condition_time.png"),
  fig_5c,
  height = 2.1,
  width = 4,
  units = "in",
  dpi = 600,
  device = ragg::agg_png,
  bg = "white"
)

fig_5c
```

# pseudobulk
```{r pseudobulk}
if (file.exists(figure5_pseudobulk_counts_path)) {
  cts.split.modified <- readRDS(figure5_pseudobulk_counts_path)
} else {
  cts <- Seurat::AggregateExpression(
    res_0.9,
    group.by = c("cell_type_fine_full", "sample"),
    slot = "counts",
    assays = "originalexp",
    return.seurat = FALSE
  )$originalexp

  cts.t <- cts |>
    as.data.frame() |>
    t() |>
    as.data.frame()

  cts.split.modified <- cts.t |>
    split(f = factor(gsub("_.*", "", rownames(cts.t)))) |>
    purrr::map(
      ~ .x |>
        `rownames<-`(str_replace(rownames(.x), "(.*)_(L\\d+)-(T\\d)", "\\2_\\3")) |>
        t()
    )

  saveRDS(cts.split.modified, figure5_pseudobulk_counts_path)
}

participants <- readr::read_csv(
  here::here("data", "metadata", "sc_sampleData.csv"),
  show_col_types = FALSE
) |>
  mutate(
    sample = as.character(sample),
    Subject_ID = factor(Subject_ID),
    time = relevel(factor(time), ref = "T4")
  ) |>
  dplyr::select(sample, Subject_ID, time)

figure5_de_samples <- c("L204127_T1", "L204127_T4", "L204185_T1", "L204185_T4")

if (file.exists(figure5_de_time_path)) {
  de_time <- readRDS(figure5_de_time_path)
} else {
  countData_list <- cts.split.modified |>
    purrr::keep(~ all(figure5_de_samples %in% colnames(.x))) |>
    purrr::map(
      ~ as.data.frame(.x) |>
        dplyr::select(dplyr::all_of(figure5_de_samples)) |>
        dplyr::filter(rowSums(.) > 0)
    )

  colData_list <- countData_list |>
    purrr::map(
      ~ participants |>
        dplyr::filter(sample %in% colnames(.x)) |>
        dplyr::arrange(match(sample, colnames(.x))) |>
        tibble::column_to_rownames("sample")
    )

  stopifnot(
    all(purrr::map2_lgl(
      countData_list,
      colData_list,
      ~ identical(colnames(.x), rownames(.y))
    ))
  )

  time <- purrr::map2(
    countData_list,
    colData_list,
    ~ DESeq2::DESeqDataSetFromMatrix(
      countData = round(.x),
      colData = .y,
      design = ~ Subject_ID + time
    ) |>
      DESeq2::DESeq()
  )

  names(time) <- names(countData_list)

  de_time <- purrr::map_dfr(
    .x = time,
    ~ DESeq2::results(.x, contrast = c("time", "T1", "T4")) |>
      as.data.frame() |>
      rownames_to_column("ID"),
    .id = "cell"
  )

  gene_annotation <- tibble(ID = unique(de_time$ID)) |>
    mutate(
      gene = AnnotationDbi::mapIds(
        org.Hs.eg.db::org.Hs.eg.db,
        keys = ID,
        keytype = "ENSEMBL",
        column = "SYMBOL",
        multiVals = "first"
      ),
      role = AnnotationDbi::mapIds(
        org.Hs.eg.db::org.Hs.eg.db,
        keys = ID,
        keytype = "ENSEMBL",
        column = "GENENAME",
        multiVals = "first"
      ),
      gene = coalesce(gene, ID),
      role = coalesce(role, "")
    )

  de_time <- de_time |>
    left_join(gene_annotation, by = "ID") |>
    dplyr::select(cell, gene, ID, role, padj, baseMean, log2FoldChange, lfcSE, stat, pvalue)

  saveRDS(de_time, figure5_de_time_path)
}
```
# 5d
### Panel D: Cell-Type Abundance Forest Plot
```{r abundance_time_glm}
data_time <- res_0.9@meta.data |>
  filter(sample %in% c("L204127_T1", "L204127_T4", "L204185_T1", "L204185_T4")) |>
  dplyr::select(cell_type_fine = cell_type_fine_full, Subject_ID, time) 

# making a list of unique cell types in the data
cell_type <- data_time |> dplyr::distinct(cell_type_fine) |> pluck(1)

# splitting the data by type of cell and 
# adding a binary variable to each split that indicates whether a row contains that cell type 
data_time_binary <- map(.x = cell_type,
                        ~ data_time |> 
                          mutate(present = case_when(cell_type_fine == .x ~ 1,
                                                     TRUE ~ 0) |> as.factor(),
                                 Subject_ID = Subject_ID |> as.factor(),
                                 time = time |> as.factor() |> relevel(ref = "T4")
                                 )
                        )

# building a glm for each cell-type split
results_time <- map_df(.x = data_time_binary, .id = "cell",
                       ~ glm(data = .x, present ~ Subject_ID + time, family = binomial) |>
                         broom::tidy(conf.int = T) |>
                         filter(term == "timeT1")# the other coef is for Subject_ID level
                         ) |> 
  mutate(cell = cell_type)

results_time2 <- results_time |>
    mutate(cell = cell |>
           str_replace(pattern = "Non-switched memory B cells", 
                                   replacement = "NSMB"),
         cell = cell |>
           str_replace(pattern = "Terminal effector CD4 T cells", 
                                   replacement = "TE CD4 T Cells"),
         cell = cell |>
           str_replace(pattern = "Myeloid dendritic cells", 
                                   replacement = "Myeloid DC"),
         cell = cell |>
           str_replace(pattern = "Follicular helper T cells", 
                                   replacement = "FH T cells")
    ) |> 
  mutate(`FDR < 0.05` = p.adjust(p.value, "fdr") <= 0.05) |>
  filter(round(p.value, 2) < 0.05) |> 
  # filter(`FDR < 0.05` == TRUE) |>
     mutate(
     estimate_exp = estimate |> exp(),
     conf.low_exp = conf.low |> exp(),
     conf.high_exp = conf.high |> exp(),
     estimate_log10 = estimate |> exp() |> log10(),
     conf.low_log10 = conf.low |> exp() |> log10(),
     conf.high_log10 = conf.high |> exp() |> log10()
                           ) |> 
  mutate(effect_size = case_when(estimate_exp > 1 ~ conf.low_exp-1,
                                 estimate_exp < 1 ~ conf.high_exp-1)) |> 
  mutate(cell = fct_reorder(cell, estimate_exp))

fig_5d <- results_time2 |> 
  ggplot(aes(y=cell, x=estimate_exp, xmin=conf.low_exp, xmax=conf.high_exp, color = `FDR < 0.05`
             )
         ) +
  geom_pointrange(shape = 20) +
  geom_segment(x = 1,
               xend = 1,
               y = 0.0,
               yend = 13,
               color = "gray") +
  expand_limits(x = c(0, 2)) +
  xlab("Odds ratio") +
  # scale_color_discrete(name = "FDR", labels = c("> 0.05", "< 0.05")) +
  scale_x_continuous(
    breaks = c(0, 0.5, 1, 1.5, 2),
    sec.axis = sec_axis(
      trans = ~.,
      breaks = c(0.25, 1, 1.75),
      labels = c("\u27f5  Enriched at T4", "No change", "Enriched at T1 \u27f6")
      )
    ) +
  geom_text(data = results_time2 |> 
              filter(estimate_exp >= 1),
            aes(label = cell, x = conf.high_exp),
            color = "black",
            hjust = - 0.01,
            family = "Arial") +
  geom_text(data = results_time2 |>
              filter(estimate_exp <= 1),
            aes(label = cell, x = conf.low_exp),
            color = "black",
            hjust = 1.01,
            family = "Arial") +
  ggthemes::theme_tufte() + 
  theme(
    text = element_text( #size=16,  
                        family = "Arial"),
    axis.text = element_text(color = "black", 
                             size = 11), # default size of tick labels is 9
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_line(color = "grey"),
    legend.position = c(0.90, 0.12),
    axis.line.x.bottom = element_line(color = "grey")
    )

 ggsave(
here::here("fig_5_files", "figure_05d_pbmc_cell_type_abundance_forest.png"),
  fig_5d,
  device = ragg::agg_png,
  width = 3.55, 
  height = 2.5, 
  units = "in", 
  res = 300,
  scaling = 0.7,
  bg = "white"
)

fig_5d
```

# 5e

### Panel E: Differentially Expressed Gene Counts by Cell Type
```{r de_number_plot}
# total number of significant genes per cell type
fig_5e <- de_time |> 
  filter(padj <= 0.05) |> 
  dplyr::count(cell) |> 
  mutate(developmental_order = # to arrange by developmental order after n
           case_when(
             cell == "Classical monocytes" ~ 1,
             cell == "Intermediate monocytes" ~ 2,
             cell == "Myeloid dendritic cells" ~ 3,
             cell == "Natural killer cells" ~ 4,
             cell == "Naive B cells" ~ 5,
             cell == "Plasmablasts" ~ 6,
             cell == "Switched memory B cells" ~ 7,
             cell == "Naive CD4 T cells" ~ 8,
             cell == "Th17 cells" ~ 9
           )
         ) |> 
  arrange(desc(n), developmental_order) |> # arranging by n, then developmental order
  rownames_to_column() |> # passing the doubly-ordered rows through to the next step 
  mutate(cell = fct_reorder(cell, rev(rowname))) |>
  ggplot(aes(x = cell, y = n)) +
    geom_segment(aes(xend = cell, yend = 0)) +
    geom_point(size = 4, color= "#e14327") +
    coord_flip() +
    xlab("") +
  ylab("Number of differentially expressed genes") +
  scale_y_continuous(expand = expansion(mult = 0, add = c(0, 2))) +
    ggthemes::theme_tufte() +
  theme(text = element_text( #size=16,  
    family = "Arial"),
    axis.text = element_text(color = "black"),
    axis.ticks.y = element_blank())
  
 ggsave(
here::here("fig_5_files", "figure_05e_pbmc_differential_gene_counts.png"),
  fig_5e,
  device = ragg::agg_png,
  width = 7.5, height = 2.4, units = "in", 
  res = 300,
  scaling = 1.2,
  bg = "white")

fig_5e
```

# 5f
### Panel F: Monocyte Pathway-Level Gene Expression
```{r monocyte_expression_pathway}
# assign pathways to genes----
pathways <- data.frame(
                     stringsAsFactors = FALSE,
                          check.names = FALSE,
                                 gene = c("PLAUR","HBEGF","EREG",
                                          "EGR1","SGK1","BCL2A1","PF4","PPBP",
                                          "ATF3","NR4A1","IER3","KLF10","MXD1",
                                          "C5AR1","NAMPT","BTG2","RGS2",
                                          "RIPK2","NFIL3","IL-8","CCL3L1",
                                          "CCL3L3","IFITM3","STK17B","CXCL8",
                                          "TMEM170B","RASGEF1B","ZFP36","GPR183"),
                       Procoagulation = c(1L,1L,0L,0L,0L,0L,1L,1L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L),
  Angiogenesis = c(1L,1L,1L,1L,1L,1L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,1L,0L,0L,
                                          0L,0L),
           `Int. stress response` = c(0L,0L,0L,1L,1L,0L,0L,0L,
                                          1L,1L,1L,1L,1L,1L,1L,1L,1L,
                                          1L,0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          1L,0L),
           `Transcription repression` = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,1L,0L,0L,0L,0L,0L,
                                          0L,1L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L),
                     `Cell survival` = c(0L,0L,0L,0L,0L,1L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,1L,1L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L),
  `Alt. mac. polarization` = c(0L,0L,0L,0L,1L,0L,0L,0L,
                                          0L,1L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L),
            `Proinflammatory` = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          1L,0L,1L,1L,1L,0L,0L,1L,0L,1L,
                                          0L,1L),
               `Interferon-inducible` = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,1L,0L,0L,0L,1L,
                                          0L,0L),
                            Apoptosis = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,1L,0L,0L,0L,
                                          1L,0L),
                  `Signal transduction` = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,1L,1L,
                                          0L,0L),
                      `mRNA regulation` = c(0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          0L,0L,0L,0L,0L,0L,0L,0L,0L,0L,
                                          1L,0L)
)
  
# Filter significant genes and ensure distinct entries
de_time_filtered <- de_time %>%
  filter(
    padj < 0.05,
    cell == "Classical monocytes",
    gene %in% pathways$gene
  )

# Sort genes by log2FoldChange
sorted_genes <- de_time_filtered %>%
  arrange(log2FoldChange) %>%
  pull(gene)

# Prepare data for heatmap
binary_intersection_matrix <- pathways %>%
  mutate(across(everything(), ~ replace_na(.x, 0))) %>%
  filter(gene %in% sorted_genes) %>%
  mutate(gene = factor(gene, levels = sorted_genes)) %>%
  pivot_longer(cols = -gene, names_to = "pathway", values_to = "membership")

binary_intersection_matrix <- binary_intersection_matrix |> 
  mutate(pathway = pathway |> factor(levels = 
      binary_intersection_matrix |> 
      filter(membership != 0) |> 
      dplyr::count(pathway) |> 
      arrange(desc(n)) |> pluck("pathway")
                                       )
      )

# Create heatmap
upset_plot <- 
  binary_intersection_matrix |> 
  ggplot(aes(x = pathway, y = gene)) +
  geom_point(aes(color = factor(membership), shape = factor(membership)), size = 2.5) +
  scale_color_manual(values = c("0" = "grey90", "1" = "gray10"), guide = "none") +
  scale_shape_manual(values = c("0" = 20, "1" = 20), guide = "none") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8, color = "black"),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    legend.position = "none",
    plot.margin = margin(0, 0, 0, 0)
  ) 

# Create gene expression plot
gene_expression_plot <- de_time_filtered %>%
  mutate(gene = factor(gene, levels = sorted_genes)) %>%
  ggplot(aes(
    x = log2FoldChange,
    y = gene,
    color = -log10(pvalue),
    xmin = log2FoldChange - 1.96 * lfcSE,
    xmax = log2FoldChange + 1.96 * lfcSE
  )) +
  geom_point(shape = 19, size = 0.75) +
  geom_linerange(size = 0.25) +
 theme(
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title.x = element_text(hjust = 0.5, vjust = 36, size = 8),
    axis.text.x = element_text(size = 8, color = "black"),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(), 
        legend.margin =  margin(0, 0, 0, 0),
        legend.position = c(0.5, -0.35),
        legend.direction = "horizontal",
        legend.title.position = "bottom",
    legend.title = element_text(hjust = 0.5, vjust = 3, size = 8),
    plot.margin = margin(0, 5, 0, 0)
  ) +
  labs(
    x = "log2(Fold Change) ± 95% CI",
    color = "-log10(p-value)"
  ) +
geom_segment(x = 0,
               y = 0,
               xend = 0,
               yend = 28,
               color = "gray") +
  scale_x_continuous(
      limits = c(-2.49, 2.49), 
      expand = c(0, 0),
      sec.axis = sec_axis(
      trans = ~.,
      breaks = c(-1.5, 0, 1.5),
      labels = c("\u27f5 Up at T4", "No change", "Up at T1 \u27f6")
      )
    ) +
      scale_color_gradient(high = "#132B43", low = "#56B1F7")  # Reverse if using blue shades

# Combine plots
fig_5f <- gene_expression_plot + upset_plot

fig_5f

# Save the plot
ggsave(
  here::here("fig_5_files", "figure_05f_monocyte_pathway_expression.png"),
  plot = fig_5f,
  device =  ragg::agg_png,
  width = 7,
  height = 3.5,
  units = "in",
  dpi = 600
  )

fig_5f
```

```{r session_info}
sessionInfo()
```
